### 2014-03-14 text/Chapter00.tex
Editing introduction chapter

\todo{This chapter is currently copy \& paste of a \texttt{}popular'' introduction to coeffects
that I wrote online (\url{http://tomasp.net/blog/2014/why-coeffects-matter}). It needs to be
rewritten in academic style - but hopefully it has the right structure for an intro chapter? }

@@

Many advances in programming language design are driven by some practical motivations.
Sometimes, the practical motivations are easy to see - for example, when they come from
some external change such as the rise of multi-core processors. Sometimes, discovering 
the practical motivations is a tricky task - perhaps because everyone is used to a 
certain way of doing things that we do not even \emph{see} how poor our current solution is.

The following two examples are related to the work done in F\# (because this is what
I'm the most familiar with), but you can surely find similar examples in other languages:

\begin{itemize}
\item \textbf{Multi-core} is an easy to see challenge caused by an external development. 
   It led to the popularity of \emph{immutable} data structures (and functional programming,
   in general) and it was also partly motivation for asynchronous workflows \cite{xx}.

\item \textbf{Data access} is a more subtle challenge. Technologies like LINQ \cite{xx} make it
   significantly easier, but it was not easy to see that inline SQL was a poor solution.
   This is even more the case for F\# \emph{type providers}. You will not realize how poor the
   established data access story is until you \emph{see} something like 
   the WorldBank and R provider \cite{xx} or CSV type provider \cite{xx}.
\end{itemize}

I believe that the next important practical challenge for programming language designers
is of the kind that is not easy to see -- because we are so used to doing things in 
certain ways that we cannot see how poor they are. The problem is designing languages
that are better at working with (and understanding) the \emph{context} in which programs are
executed.

@@

\section{Context-aware programming matters}

The phrase \emph{context in which programs are executed} sounds quite abstract and generic.
What are some concrete examples of such \emph{context}? For example:

\begin{itemize}
\item When writing a cross-platform application, different platforms (and even different 
   versions of the same platform) provide different contexts - the API functions that are 
   available.

\item When creating a mobile app, the different capabilities that you may (or may not) have
   access to are context (GPS sensor, accelerometer, battery status).

\item When working with data (be it sensitive database or social network data from Facebook),
   you have permissions to access only some of the data (depending on your identity) and
   you may want to track \emph{provenance} information. This is another example of a context.
\end{itemize}

These are all fairly standard problems that developers deal with today. As the number of 
devices where programs need to run increases, dealing with diverse contexts will be becoming
more and more important (and I'm not even talking about \emph{ubiquitous computing} where you need
to compile your code to a coffee machine).

We do not preceive the above things as problems (at best, annoyances that we just have to 
deal with), because we do not realize that there should be a better way. Let me digg into
four examples in a bit more detail.

@@

The number of different platform versions that you need to target is increasing, no matter what 
platform are you using. For Android, there is a number called API Level 
\footnote{\url{http://developer.android.com/guide/topics/manifest/uses-sdk-element.html}}
with mostly additive changes (\emph{mostly} sounds very reassuring). In the .NET world, there are multiple
versions, mobile editions, multiple versions of Silverlight etc. So, code that targets multiple
frameworks easily ends up looking as the following sample from the Http module in F\# Data\footnote{
\url{https://github.com/fsharp/FSharp.Data/blob/master/src/Net/Http.fs}}:

\begin{verbatim}
for header, value in headers do
  match header with
  | StringEquals "accept" -> req.Accept <- value
#if FX_NO_WEBREQUEST_USERAGENT
  | StringEquals "user-agent" -> 
      req.Headers.[HttpRequestHeader.UserAgent] <- value
#else
  | StringEquals "user-agent" -> req.UserAgent <- value
#endif
#if FX_NO_WEBREQUEST_REFERER
  | StringEquals "referer" -> 
      req.Headers.[HttpRequestHeader.Referer] <- value
#else
  | StringEquals "referer" -> req.Referer <- value
#endif
  | _ -> req.Headers.[header] <- value
\end{verbatim}

This is difficult to write (you won't know if the code even compiles until you try building all 
combinations) and maintaining such code is not particularly nice (try adding another platform version).
Maybe we could refactor the code in F\# Data to improve it, but that's not really my point - supporting
multiple platforms should be a lot easier.

On Android, you can access API from higher level platform dynamically using techniques like 
reflection and writing wrappers 
\footnote{\url{http://android-developers.blogspot.cz/2009/04/backward-compatibility-for-android.html}}
Again, this sounds very error prone - to quote the linked article 
\emph{``Remember the mantra: if you haven't tried it, it doesn't work''}.
Testing is no doubt important. But at least in statically typed langauges, we
should not need to worry about calling a method that does not exist when writing
multi-platform applications!

@@

\subsubsection{Context awareness \#2: System capabilities}

Another example related to the previous one is when you're using something like LINQ or 
FunScript (\url{http://funscript.info/}) to translate code written in a sub-set of C\# or F\#
to some other language - such as SQL or JavaScript. This is another important area, because
you can use this technique to target JavaScript, GPUs, SQL, Erlang runtime or other components 
that use a particular programming language.

For example, take a look at the following LINQ query that selects product names where the
first upper case letter is "C":

\begin{verbatim}
var db = new NorthwindDataContext();

from p in db.Products
where p.ProductName.First(c => Char.IsUpper(c)) == "C"
select p.ProductName;
\end{verbatim}

This looks like a perfectly fine piece of code and it compiles fine. But when you try running
it, you get the following error:

\begin{quote}
Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators 
not supported for type \texttt{System.String}.
\end{quote}

The problem is that LINQ can only translate a \emph{subset} of normal C\# code. Here, we are using
the \texttt{First} method to iterate over characters of a string and that's not supported. This is
not a technical limitation of LINQ, but a fundamental problem of the approach. If we target a 
limited language, we simply cannot support the full source language. Is this an important 
problem? If you search on Google for ``has no supported translation to SQL'' 
(which is a part of a similar error message that you get in another case), you get some 26900
links. So yes - this is an issue that people are hitting all the time.

@@

\subsubsection{Context awareness \#3: Confidentiality and provenance}

The previous two examples were mainly related to the (non-)existence of some API functions
or to their behaviour. However, this is not the only case of context-awarness that is important
in every day programming.

Most readers of this blog will immediately see what is wrong with the following code, but that
does not change the fact that it can be compiled and deployed (and that there is a large number
of systems that contain something similar):

\begin{verbatim}
let query = sprintf "SELECT * FROM Products WHERE ProductName = '%s'" name
use cmd = new SqlCommand(query)
use reader = cmd.ExecuteReader()
\end{verbatim}

The problem is obviously SQL Injenction \footnote{\url{http://en.wikipedia.org/wiki/SQL_injection}}. 
Concatenating strings to build an SQL command is a bad practice, but it is an example of a more general problem.

Sometimes, we have special kinds of variables that should have some contextual meta-data associated
with them. Such meta-data can dictate how the variable can be used. Here, \texttt{name} comes from the user
input and this \emph{provenance} information should propagate to \texttt{query} and we should get an error when
we try passing this - potentially unsafe - input to \texttt{SqlCommand}. Similarly, if you have \texttt{password} or
\texttt{creditCard}, it should be annotated with meta-data saying that this is a sensitive piece of data and
should not, for example, be sent over an unsecured network connection.
As a side note - this idea is related to a technique called taint checking \cite{xx}.

In another context, if you are working with data (e.g. in some data jouranlism application), it would be
great if your sources were annotated with meta-data about the quality and source of the data (e.g.
can it be trusted? how up-to-date is it?) The meta-data could propagate to the result and tell you how
accurate and trustworthy your result is.

@@

\subsubsection{Context-awareness \#4: Resource \& data availability}

>>>>

Many advances in programming language design are driven by some practical motivations. Sometimes, 
the practical motivations are easy to see -- for example, when they come from an external change 
such as the rise of multi-core processors. Sometimes, discovering the practical motivations is a 
difficult task -- perhaps because we are so used to a certain way of doing things that we do not 
even \emph{see} the flaws of our approach.

@@

\section{Programming language and innovation}

We briefly consider two practical concerns that led to the development of new programming languages. 
This helps to explain why context-aware programming is of importance, which we cover in the next section. 
The examples are by no means representative, but they illustrate the motivations well.

\paragraph{Parallel programming.}

The rise of multi-core CPUs is a clear example of an external development influencing
programming language research. As multi-core and multi-processor systems became de-facto standard, 
languages had to provide better abstractions for parallel programming. This led to the industrial 
popularity of \emph{immutable} data structures (and functional programming in general), software 
transactional memory \cite{app-haskell-stm}, data-parallelism and also asynchronous 
computing \cite{app-fsharp-async}.

In this case, the motivation is easy to see -- writing multi-core programs using earlier 
abstractions, such as threads and locks, is difficult and error-prone. At the same time, 
multi-core CPUs become the standard very quickly and so the lack of good language abstractions
was apparent.

\paragraph{Data access.}

Accessing data is an example of a more subtle challenge. Initiatives like open government data
certainly make more data available. However, to access the data, one has to parse CSV and Excel files, 
issue SQL or SPARQL queries (to query database and the semantic web, respectively).

Technologies like LINQ \cite{app-linq} make querying data significantly easier. But perhaps 
because accessing data became important more gradually, it was not easy to see that inline
SQL is a poor solution.

This is even more the case for \emph{type providers} -- a recent feature in F\# that integrates
external data sources directly into the type system of the language and thus makes data explorable
directly from the source code editor (through features such as auto-completion on object members).
It is not easy to see the limitations of standard techniques (using HTTP requests to query REST 
services or parsing CSV files and using string-based lookup) until one sees how type providers 
change the data-scientist's workflow\footnote{This is difficult to explain in writing and so the 
reader is encouraged to watch a video showing type providers for the WorldBank and CSV data 
sources \cite{app-fsharp-world}.}.

\paragraph{Context-aware programming.}

In this chapter, we argue that the next important practical challenge for programming language 
designers is designing languages that are better at working with (and understanding) the 
\emph{context in which programs are executed}.

This challenge is of the kind that is not easy to see -- perhaps because we are so used to 
doing things in certain ways that we cannot see their flaws. In this chapter, we aim to
uncover such flaws -- we look at a number of basic programs that rely on contextual 
information, we explain why they are inappropriate and then we briefly outline how this
thesis remedies the situation.

Putting deeper philosophical questions about the nature of scientific prog\-ress aside,
the goal of programming language research is generally to design languages that provide 
more \emph{appropriate abstractions} for capturing common problems, are \emph{simple} and 
more \emph{unified}. 
These are exactly the aims that we follow in this thesis. In this chapter, we explain what
the common problems are. In Chapter~\ref{ch:flat-coeffects} and Chapter~\ref{ch:structural-coeffects},
we develop two simple calculi to understand and capture the structure of the problems and,
finally, Chapter~\ref{ch:unified-coeffects} unifies the two abstractions.

@@

\section{Why context-aware programming matters}

The phrase \emph{context in which programs are executed} sounds rather abstract and generic.
What notions of \emph{context} can be identified in modern software systems?
Different environments provide different resources (\eg~database
or GPS sensor), environments are increasingly diverse (\eg~multiple versions of different
mobile platforms). Web applications are split between client, server and mobile components; mobile
applications must be aware of the physical environment while the ``internet
of things`` makes the environments even more heterogeneous. At the same time, applications access
rich data sources and need to be aware of security policies and provenance information from the
environment.

Writing such context-aware (or environment-aware) applications is a fundamental problem of modern 
software engineering. The state of the art relies on ad-hoc approaches -- using hand-written conditions 
or pre-processors for conditional compilation. Common problems that developers face include:

\begin{itemize}
\item \textbf{System capabilities.} When writing code that is cross-compiled to multiple 
  targets (e.g. SQL \cite{app-linq}, OpenCL or JavaScript \cite{app-hop-lang}) a part of the compilation 
  (\eg~SQL generation) often occurs at runtime and developers have no guarantee that it will succeed 
  until the program is executed.

\item \textbf{Platform versions.} When developing cross-platform applications, different platforms
  (and different versions of the same platform) provide different API functions. Writing a 
  cross-platform code usually relies on (fragile) conditional compilation or (equally fragile) 
  dynamic loading.

\item \textbf{Security and provenance.} When working with data (be it sensitive database or social 
   network data), we have permissions to access only some of the data and we may want to track 
   \emph{provenance} information. However, this is not checked -- if a program attempts to access 
   unavailable data, the access will be refused at run-time.

\item \textbf{Resources \& data availability.} When creating a mobile application, the program may
  (or may not) be granted access to device capabilities such as GPS sensor, social updates or battery
  status. We would like to know which of the capabilities are required and which are optional
  (\ie~enhance the user experience, but there is a fallback strategy). Equally, on the server-side,
  we might have access to different database tables and other information sources.
\end{itemize}

Most developers do not perceive the above as programming language flaws -- they are simply common
programming problems (at most somewhat annoying and tedious) that had to be solved. However, this 
is because we do not realize that a suitable language extension could make the above problems 
significantly easier to solve. As the number of distinct contexts and their diversity increases,
these problems will become even more commonplace.

The following sub-sections explore four examples in more details. The examples are chosen to 
demonstrate two distinct forms of contexts that are studied in this thesis.

\begin{figure}
\begin{equation*}
\begin{array}{l}
\kvd{for}~\ident{header},\ident{value}~\kvd{in}~\ident{header}~\kvd{do}\\[-0.25em]
\hspace{2em}\kvd{match}~\ident{header}~\kvd{with}\\[-0.25em]
\hspace{2em}|~\str{accept} \rightarrow~ \ident{req.Accept} \leftarrow \ident{value}\\[-0.25em]
%
\prepk{\#if}~\prepi{FX\_NO\_WEBREQUEST\_USERAGENT}\\[-0.25em]
\hspace{2em}|~\str{user-agent} \rightarrow~ \ident{req.UserAgent} \leftarrow \ident{value}\\[-0.25em]
\prepk{\#else}\\[-0.25em]
\hspace{2em}|~\str{user-agent} \rightarrow~
  \ident{req.Headers}.[\,\ident{HttpHeader.UserAgent}\,] \leftarrow \ident{value}\\[-0.25em]
\prepk{\#endif}\\[-0.25em]
%
\prepk{\#if}~\prepi{FX\_NO\_WEBREQUEST\_REFERER}\\[-0.25em]
\hspace{2em}|~\str{referer} \rightarrow~ \ident{req.Referer} \leftarrow \ident{value}\\[-0.25em]
\prepk{\#else}\\[-0.25em]
\hspace{2em}|~\str{user-agent} \rightarrow~
  \ident{req.Headers}.[\,\ident{HttpHeader.Referer}\,] \leftarrow \ident{value}\\[-0.25em]
\prepk{\#endif}\\[-0.25em]
%
\hspace{2em}|~\ident{other} \rightarrow~ \ident{req.Headers}.[\,\ident{other}\,] \leftarrow \ident{value}\\[-0.25em]
\end{array}
\end{equation*}

\vspace{1em}
\noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}} 
\vspace{-1.5em}
\caption{Conditional compilation in the HTTP module of the F\# Data library}
\label{fig:introduction-context-http}
\end{figure}

@@

The diversity across devices means that developers need to target an increasing number of platforms
and possibly also multiple versions of each platform. For Android, there is a number called 
API level \cite{app-android-apilevel} which ``uniquely identifies the framework API revision offered 
by a version of the Android platform''. Most changes in the libraries (but not all) are additive.

Equally, in the .NET ecosystem, there are multiple versions of the .NET runtime, mobile and 
portable versions of the framework etc. The differences may be subtle -- for example, some members
are omitted to make the mobile version of the library smaller, some functionality is not available
at all, but naming can also vary between versions.

For example, the Figure~\ref{fig:introduction-context-http} shows an excerpt from the \ident{Http} module 
in the F\# Data library\footnote{The file version shown here is available at:
\url{https://github.com/fsharp/FSharp.Data/blob/b4c58f4015a63bb9f8bb4449ab93853b90f93790/src/Net/Http.fs}}.
The example uses conditional compilation to target multiple versions of the .NET framework. Such code
is difficult to write -- to see whether a change is correct, it had to be recompiled for all combinations
of pre-processor flags -- and maintaining the code is equally hard. The above example could be refactored
and the .NET API could be cleaner, but the fundamental issue remains. If the language does not understand
the context (here, the different platforms and platform versions), it cannot provide any static
guarantees about the code.

As an alternative to conditional compilation, developers can use dynamic loading. For example, 
on Android, programs can access API from higher level platform dynamically using techniques like 
reflection and writing wrappers. This is even more error prone. As noted in an article introducing
the technique\footnote{Retrieved from: \url{http://android-developers.blogspot.com/2009/04/backward-}\\\url{compatibility-for-android.html}}:
\emph{``Remember the mantra: if you haven't tried it, it doesn't work''}. Again, it would be 
reasonable to expect that statically-typed languages could provide a better solution.

@@

\subsection{Context awareness \#2: System capabilities}

Another example related to the previous one is when libraries use meta-programming techniques
(such as LINQ \cite{app-linq} or F\# quotations \cite{app-fsharp-metaprog}) to translate code
written in a subset of a host language to some other target language, such as SQL, OpenCL or 
JavaScript. This is an important technique, because it lets developers targets multiple 
heterogeneous runtimes that have limited execution capabilities.

For example, consider the following LINQ query written in C\# that queries a database and
selects product names where the first upper case letter is "C":
%
\begin{equation*}
\begin{array}{l}
\kvd{var}~\ident{db} = \kvd{new}~\ident{NorthwindDataContext}();\\[0.25em]
\kvd{from}~\ident{p}~\kvd{in}~\ident{db.Products}\\[-0.25em]
\kvd{where}~\ident{p.ProductName.First}(\ident{c} \Rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
\kvd{select}~\ident{p.ProductName};
\end{array}
\end{equation*}
%
This appears as a perfectly valid code ant the C\# compiler accepts it. However, when the program
is executed, it fails with the following error:
%
\begin{quote}
Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators 
not supported for type \texttt{System.String}.
\end{quote}
%
The problem is that LINQ can only translate a \emph{subset} of normal C\# code. The above snippet
uses \ident{First} method to iterate over characters of a string, which is not supported. This is
not a technical limitation of LINQ, but a fundamental problem of the approach.

When cross-compiling to a limited environment, we cannot always support the full source language.
The example with LINQ and SQL demonstrates the importance of this problem. As of March 2014,
Google search returns 11800 results for the message above and even more (44100 results) for a
LINQ error message \emph{``Method X has no supported translation to SQL``} caused by a similar 
limitation.

@@

\subsection{Context awareness \#3: Confidentiality and provenance}

The previous two examples were related to the non-existence of some library functions in another
environment. Another common factor was that they were related to the execution context of the whole 
program or a scope. However, contextual properties can be also related to specific variables.

For example, consider the following code sample that accesses database by building a SQL query
using string concatenation:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{query} = \ident{sprintf}~\str{SELECT * FROM Products WHERE Name='\%s'}~\ident{name}\\[-0.25em]
\kvd{let}~\ident{cmd} = \kvd{new}~\ident{SqlCommand}(\ident{query})\\[-0.25em]
\kvd{let}~\ident{reader} = \ident{cmd.ExecuteReader}()
\end{array}
\end{equation*}
%
The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
(an attacker could enter \str{'; DROP TABLE Products -{}-} as their \ident{name} and delete the 
database table with products). For this reason, most libraries discourage building SQL commands by
string concatenation, but there are still many systems that do so.

The example demonstrates a more general property. Sometimes, it is desirable to track additional 
meta-data about variables that are in some ways special. Such meta-data can determine how the variables
can be used. Here, \ident{name} comes from the user input. This \emph{provenance} information should 
be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
not directly contain user input (in an unchecked form). Such marking of values (but at run-time)
is also called tainting \cite{app-tainting-sql}.

Similarly, if we had \ident{password} or \ident{creditCard} variables in a client/server web application,
these should be annotated as sensitive and it should not be possible to send their values over an
unsecured network connection.

In another context, when working with data (\eg~in data journalism), it would be desirable to 
track meta-data about the quality and the source of the data. For example, is the source trustworthy?
Is the data up-to-date? Such meta-data could propagate to the result and tell us important 
information about the calculated results.

@@

\subsection{Context-awareness \#4: Access patterns in data-flow languages}

As an example, consider a language that allows us to get a value of a variable (representing
some changing data-source) \ident{x} versions back using the syntax $\ident{x}_{[a]}$. 
To track information about individual variables, we use a product-like operation $\times$ on tags 
to mirrors the product structure of variables. For example:
%
\begin{equation*}
?
\end{equation*}
%
The annotations $5 \times 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
that we need at most 5 and 10 past values of \ident{a} and \ident{b}, respectively. If we substitute 
\ident{c} for both \ident{a} and \ident{b}, we need to combine the context-requirements and take the
maximum of the requirements of the individual variables:
%
\begin{equation*}
\begin{array}{l}
?
\end{array}
\end{equation*}
%
This version of the calculus removes the non-determinism of the lambda abstraction from the previous
version. As we associate information with individual variables, lambda abstraction creates a function
that requires the context requirements associated with the variable that is being abstracted over.

For example, if we wrapped the earlier example above in a function taking \ident{a} (and using 
\ident{b} from the declaration-site) then the function would have context requirements $5$ -- that
is the number associated with the variable \ident{a}.

\subsection{Context-awareness \#4: Resource \& data availability}
